A Tutorial on Tailored Simulation-Based Power Analysis for Experimental Designs with Generalized Linear Mixed Models

### 2024-04-08 docs/manuscript.Rmd
fix escape _ in \text

In such a setting, we model this probability as
$$
P_{si} = \text{inverse_logit}(\eta_{si})
$$
with the inverse-logit link $g^{-1}(\eta_{si}) = inverse\_logit(\eta_{si}) = \frac{exp(\eta_{si})}{1 + exp(\eta_{si})}$ or equivalently
$$
\text{logit}(P_{si}) = \eta_{si}
$$
with the logit link $g(P_{si}) = \text{logit}(P_{si}) = \text{ln} (\frac{P_{si}}{1 - P_{si}})$.

@@

In total, we assume the model
$$
\begin{aligned}
\text{logit}[P_{si}] =\ (&\beta_0 + u_{0s} + u_{0i}) + \\
&\beta_a \cdot advice\_present_{si} + \beta_c \cdot advice\_correct_{si} + \beta_e \cdot expert_s + \\
&\beta_{ea} \cdot expert_{s} \cdot advice\_present_{si} + \beta_{ec} \cdot expert_{s} \cdot advice\_correct_{si}
\end{aligned}
$$
or equivalently
$$
\begin{aligned}
P_{si} = \text{inverse_logit}[&(\beta_0 + u_{0s} + u_{0i}) + \\
&\beta_a \cdot advice\_present_{si} + \beta_c \cdot advice\_correct_{si} + \beta_e \cdot expert_s + \\
&\beta_{ea} \cdot expert_{s} \cdot advice\_present_{si} + \beta_{ec} \cdot expert_{s} \cdot advice\_correct_{si}]
\end{aligned}
$$
with model parameters $\beta_0$, $\beta_e$, $\beta_a$, $\beta_c$, $\beta_{ea}$, $\beta_{ec}$, $\sigma_S$, and $\sigma_I$.

@@

<!-- $$ -->
<!-- \begin{aligned} -->
<!-- & \text{inverse_logit}[\beta_0 + \beta_a \cdot 1 + \beta_c \cdot 1 + \beta_e \cdot 1 + \beta_{ea} \cdot 1 \cdot 1 + \beta_{ec} \cdot 1 \cdot 1]\\ -->
<!-- & \quad - \text{inverse_logit}[\beta_0 + \beta_a \cdot 0 + \beta_c \cdot 0 + \beta_e \cdot 1 + \beta_{ea} \cdot 1 \cdot 0 + \beta_{ec} \cdot 1 \cdot 0] = \\ -->
<!-- & = \text{inverse_logit}[\beta_0 + \beta_a  + \beta_c + \beta_e + \beta_{ea} + \beta_{ec} ]\\ -->
<!-- & \quad - \text{inverse_logit}[\beta_0 + \beta_e ] \\ -->
<!-- \end{aligned} -->
<!-- $$ -->

@@

Unfortunately, there is no single model parameter that corresponds with our contrast of interest.
the model parameters in a binomial GLMM are hard to interpret because 1) the $\beta$ parameters are connected to the modeled probability via the non-linear inverse-logit link, and 2) we also have to consider the random effects.
When we are interested in a subject with average ability ($u_{0s} = 0$) responding to an item (i.e., CT scan) with average difficulty ($u_{0i} = 0$), the model implied probability that such a person solves such an item accurately is given by: 
$$
\begin{aligned}
P(Y=1|\mathbf{X=x}, \mathbf{U} = \mathbf{0}) = \\
= \text{inverse_logit}[&\beta_0 + \beta_a \cdot advice\_present_{si} + \beta_c \cdot advice\_correct_{si} + \beta_e \cdot expert_s + \\
&\beta_{ea} \cdot expert_{s} \cdot advice\_present_{si} + \beta_{ec} \cdot expert_{s} \cdot advice\_correct_{si}]
\end{aligned}
$$
In fact, we would only need the full equation if the subject is an expert and correct advice is presented.
In all other experimental conditions, some terms drop from the equation because they are multiplied by $0$.
The other extreme case would be the probability that a non-expert with average ability solves an item with average difficulty when no advice is presented:
$$
\begin{aligned}
P(Y=1| advice\_present = 0, advice\_correct = 0, expert = 0, u_{0s} = 0, u_{0i} = 0) = \\
= \text{inverse_logit}[\beta_0]
\end{aligned}
$$
One could use such reasoning, to determine the exact formula needed to compute our contrast of interest.
However, this would be very tedious and we can use R to compute this contrast for us without doing the math.

>>>>

In such a setting, we model this probability as
$$
P_{si} = \text{inverse\_logit}(\eta_{si})
$$
with the inverse-logit link $g^{-1}(\eta_{si}) = inverse\_logit(\eta_{si}) = \frac{exp(\eta_{si})}{1 + exp(\eta_{si})}$ or equivalently
$$
\text{logit}(P_{si}) = \eta_{si}
$$
with the logit link $g(P_{si}) = \text{logit}(P_{si}) = \text{ln} (\frac{P_{si}}{1 - P_{si}})$.

@@

In total, we assume the model
$$
\begin{aligned}
\text{logit}[P_{si}] =\ (&\beta_0 + u_{0s} + u_{0i}) + \\
&\beta_a \cdot advice\_present_{si} + \beta_c \cdot advice\_correct_{si} + \beta_e \cdot expert_s + \\
&\beta_{ea} \cdot expert_{s} \cdot advice\_present_{si} + \beta_{ec} \cdot expert_{s} \cdot advice\_correct_{si}
\end{aligned}
$$
or equivalently
$$
\begin{aligned}
P_{si} = \text{inverse\_logit}[&(\beta_0 + u_{0s} + u_{0i}) + \\
&\beta_a \cdot advice\_present_{si} + \beta_c \cdot advice\_correct_{si} + \beta_e \cdot expert_s + \\
&\beta_{ea} \cdot expert_{s} \cdot advice\_present_{si} + \beta_{ec} \cdot expert_{s} \cdot advice\_correct_{si}]
\end{aligned}
$$
with model parameters $\beta_0$, $\beta_e$, $\beta_a$, $\beta_c$, $\beta_{ea}$, $\beta_{ec}$, $\sigma_S$, and $\sigma_I$.

@@

<!-- $$ -->
<!-- \begin{aligned} -->
<!-- & \text{omv}[\beta_0 + \beta_a \cdot 1 + \beta_c \cdot 1 + \beta_e \cdot 1 + \beta_{ea} \cdot 1 \cdot 1 + \beta_{ec} \cdot 1 \cdot 1]\\ -->
<!-- & \quad - \text{inverse_logit}[\beta_0 + \beta_a \cdot 0 + \beta_c \cdot 0 + \beta_e \cdot 1 + \beta_{ea} \cdot 1 \cdot 0 + \beta_{ec} \cdot 1 \cdot 0] = \\ -->
<!-- & = \text{inverse_logit}[\beta_0 + \beta_a  + \beta_c + \beta_e + \beta_{ea} + \beta_{ec} ]\\ -->
<!-- & \quad - \text{inverse_logit}[\beta_0 + \beta_e ] \\ -->
<!-- \end{aligned} -->
<!-- $$ -->

@@

Unfortunately, there is no single model parameter that corresponds with our contrast of interest.
the model parameters in a binomial GLMM are hard to interpret because 1) the $\beta$ parameters are connected to the modeled probability via the non-linear inverse-logit link, and 2) we also have to consider the random effects.
When we are interested in a subject with average ability ($u_{0s} = 0$) responding to an item (i.e., CT scan) with average difficulty ($u_{0i} = 0$), the model implied probability that such a person solves such an item accurately is given by: 
$$
\begin{aligned}
P(Y=1|\mathbf{X=x}, \mathbf{U} = \mathbf{0}) = \\
= \text{inverse\_logit}[&\beta_0 + \beta_a \cdot advice\_present_{si} + \beta_c \cdot advice\_correct_{si} + \beta_e \cdot expert_s + \\
&\beta_{ea} \cdot expert_{s} \cdot advice\_present_{si} + \beta_{ec} \cdot expert_{s} \cdot advice\_correct_{si}]
\end{aligned}
$$
In fact, we would only need the full equation if the subject is an expert and correct advice is presented.
In all other experimental conditions, some terms drop from the equation because they are multiplied by $0$.
The other extreme case would be the probability that a non-expert with average ability solves an item with average difficulty when no advice is presented:
$$
\begin{aligned}
P(Y=1| advice\_present = 0, advice\_correct = 0, expert = 0, u_{0s} = 0, u_{0i} = 0) = \\
= \text{inverse\_logit}[\beta_0]
\end{aligned}
$$
One could use such reasoning, to determine the exact formula needed to compute our contrast of interest.
However, this would be very tedious and we can use R to compute this contrast for us without doing the math.
